Fast Sorting on Modern Computers

### 2005-08-01 docs/tech_report/heap.tex
added testing comments
started testing

%TODO move this to testing section
Figure \ref{heap cycles} shows the actual time taken for heap sort on a Pentium
4 1.8GHz machine, which has an 8-way associative 256KB level 2 cache with a 64 byte
cache line, a 4-way associative 8KB level 1 data cache, and a maximum branch
misprediction delay of 20 cycles\footnote{This is determined from the Pentium
4's 20 stage pipeline. This number is not exact, and is probably slightly lower,
with the possibility of occasionally rising higher (in the case of an
instruction cache miss, which is exceptionally unlikely for small routines like
sorts).}. Its static branch prediction strategy is backward taken and forward
not taken; its dynamic strategy is unspecified: though \cite{Intel248966-010}
mentions a branch history so it is likely a two-level adaptive predictor.

The figure shows that the base heapsort performs best when the data set fits in
the cache. After that point, the 8-heap performs best, and its cycle count increases
with a smaller slope. This shows the importance of LaMarca's cache-conscious
design even nine years after his thesis. It also shows the importance of reduced
branch mispredictions: the base heapsort has more instructions and level 1 cache
misses, but still outperforms the 4-heap, until level 2 cache misses take over.
However, this is hardly conclusive: the lower level 2 cache misses may account
for this.

>>>>

Figure \ref{heap cycles} shows the actual time taken for heap sort on a Pentium
4 machine.  The figure shows that the base heapsort performs best when the data
set fits in the cache. After that point, the 8-heap performs best, and its cycle
count increases with a smaller slope. This shows the importance of LaMarca's
cache-conscious design even nine years after his thesis. It also shows the
importance of reduced branch mispredictions: the base heapsort has more
instructions and level 1 cache misses, but still outperforms the 4-heap, until
level 2 cache misses take over.  However, this is hardly conclusive: the lower
level 2 cache misses may account for this.

### 2005-08-01 docs/tech_report/method.tex
started testing

SimpleScalar works by compiling code into its own instruction set. It provides a
compiler - a port of gcc 2.63 - with related tools such as a linker and
assembler.  Several simulators are then provided: \cc{sim-fast} only counts
instruction, quickly; \cc{sim-cache} simulates a cache and counts level 1 and
level 2 cache misses; \cc{sim-bpred} simulates a branch predictor and counts
branch hits and misses. Other simulators are provided, but they were not used in
our tests.

@@

%TESTING - also what version
The SimpleScalar settings were chosen to be similar to LaMarca's simulations.
LaMarca used the \n{Atom} simulator, which was used to test software for the DEC
AlphaStation. It simulated a 2MB cache with a 32 byte cache line. The remainder
of the settings were taken from the DEC Alpha: 8KB level 1 data cache, 8KB level
1 instruction cache and shared instruction and data level 2 cache.

%TESTING - what versino
\subsection{Gnuplot}
\label{Gnuplot}
\cc{Gnuplot} is a tool for plotting data files. It provides visualisations of data in
two and three dimensions, and is used to create all of the graphs in this
document. Its homepage is \n{http://www.gnuplot.info/}.

@@

%TESTING - what version
\subsection{Valgrind}
\label{Valgrind}
\cc{Valgrind} is a set of tools which are used for debugging and profiling.  By
default it works as a memory checker, and detects errors such as the use of
uninitialised memory and accessing memory off the end of an array. This use
makes it particularly useful for bounds checking. No configuration was required
for this, and simply running \cc{valgrind a.out} ran the program through
valgrind. The homepage for valgrind is at \n{http://valgrind.kde.org/}.

%TESTING - what version
\subsection{PapiEx}
\cc{PapiEx} is a tool used to access performance counters on a Pentium 4. It
uses the PAPI framework to count certain characteristics across a program's
execution. We used it to count the number of cycles a sort took to run. This
provided an accurate measure of how fast a sort executes on a Pentium 4. The
PapiEx homepage is at \n{http://icl.cs.utk.edu/~mucci/papiex/}.

@@

>>>>

SimpleScalar works by compiling code into its own instruction set. It provides a
compiler - a port of gcc - with related tools such as a linker and assembler.
Several simulators are then provided: \cc{sim-fast} only counts instruction,
quickly; \cc{sim-cache} simulates a cache and counts level 1 and level 2 cache
misses; \cc{sim-bpred} simulates a branch predictor and counts branch hits and
misses. Other simulators are provided, but they were not used in our tests.

@@

\subsection{Gnuplot}
\label{Gnuplot}
\cc{Gnuplot} is a tool for plotting data files. It provides visualisations of
data in two and three dimensions, and is used to create all of the graphs in
this document. Its homepage is \n{http://www.gnuplot.info/}.

@@

\subsection{Valgrind}
\label{Valgrind}
\cc{Valgrind} is a set of tools which are used for debugging and profiling.  By
default it works as a memory checker, and detects errors such as the use of
uninitialised memory and accessing memory off the end of an array. This use
makes it particularly useful for bounds checking. No configuration was required
for this, and simply running \cc{valgrind a.out} ran the program through
valgrind. The homepage for valgrind is at \n{http://valgrind.kde.org/}.

\subsection{PapiEx}
\cc{PapiEx} is a tool used to access performance counters on a Pentium 4. It
uses the PAPI framework to count certain characteristics across a program's
execution. We used it to count the number of cycles a sort took to run. This
provided an accurate measure of how fast a sort executes on a Pentium 4. The
PapiEx homepage is at \n{http://icl.cs.utk.edu/~mucci/papiex/}.

@@

\subsection{}
simplescalar version and settings used
The SimpleScalar settings were chosen to be similar to LaMarca's simulations.
LaMarca used the \n{Atom} simulator, which was used to test software for the DEC
AlphaStation. It simulated a 2MB cache with a 32 byte cache line. The remainder
of the settings were taken from the DEC Alpha: 8KB level 1 data cache, 8KB level
1 instruction cache and shared instruction and data level 2 cache.

\gnuplot version
valgrinf version
papiex version

cache line length. - vs pentium 4

penitum 4:
1.8, 8-way 256K level 2 64 byte line (note i wrote 128 byte xsomewhere. is this
right)
4-way 8kb level 1 cache

branch delay 20 cycles

1.8GHz machine, which has an 8-way associative 256KB level 2 cache with a 64 byte
cache line, a 4-way associative 8KB level 1 data cache, and a maximum branch
misprediction delay of 20 cycles\footnote{This is determined from the Pentium
4's 20 stage pipeline. This number is not exact, and is probably slightly lower,
with the possibility of occasionally rising higher (in the case of an
instruction cache miss, which is exceptionally unlikely for small routines like
sorts).}. Its static branch prediction strategy is backward taken and forward
not taken; its dynamic strategy is unspecified: though \cite{Intel248966-010}
mentions a branch history so it is likely a two-level adaptive predictor.

64 bit vs 32 bit
